getdata4.Rmd

### 2022-02-18 getdata.Rmd
modifications ...
clean complete, fresh start

---
title: "Get population and temperature data"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(eha)
library(skum)
options(show.signif.stars = FALSE)
g <- function(x) (log(1 + 365 * x))^3
```

# Population data

We use the observation file rather than the  population file, so we can get 
correct parish information.

```{r readobs}
bb <- obs$id[obs$starttyp == 2]
births <- obs[obs$id %in% bb, 
    c("id", "sex", "birthdate", "nofrs", "sluttyp", "socStatus", "hisclass", 
      "enter", "exit", "ort", "ortnmn")]
births$event <- as.numeric(births$sluttyp == 2)
births <- age.window(births, c(0, 1))
births <- births[order(births$id, births$enter), ]
##indx <- tapply(births$id, births$id)
##births$mexit <- tapply(births$exit, births$id, max)[indx]
##births$mevent <- tapply(births$event, births$id, max)[indx]
##births <- births[!duplicated(births$id), ]
### Deaths:
did <- unique(births$id[births$event == 1])
deaths <- births[births$id %in% did, ]
indx <- tapply(deaths$id, deaths$id)
deaths$menter <- tapply(deaths$enter, deaths$id, min)[indx]
deaths <- deaths[!duplicated(deaths$id), ]
deaths$enter <- deaths$menter
deaths$menter <- NULL
deaths$enter <- 0 ## satisfied....
### Survivors:
survs <- births[!(births$id %in% did), ]
indx <- tapply(survs$id, survs$id)
survs$mexit <- tapply(survs$exit, survs$id, max)[indx]
survs <- survs[!duplicated(survs$id), ]
survs$exit <- survs$mexit
survs$mexit <- NULL
survs$enter <- 0 ## Ok ...
### Now, give birth to NEW births:
births <- rbind(deaths, survs)
births <- births[births$birthdate > 1895, ]

indx <- match(births$id, pers$id)
births$ab <- pers$ab[indx]
##births$df <- pers$df[indx]
births$illeg <- factor(births$ab == 2, labels = c("no", "yes"))
births$ab <- NULL
births$sluttyp <- NULL
births$parity <- pers$paritet_g[indx]
summary(births)                                
```

We put on 'subreg':

```{r subreg}
births$subreg <- factor(births$nofrs)
levels(births$subreg) <- c(rep("ume", 2), rep("bjur", 4), rep("sten", 3))
## Remove "sten":
births <- births[births$subreg != "sten", ]
births$subreg <- factor(births$subreg) # Remove level "sten"
```

# Neonatal mortality, an overview

```{r neo, fig.cap = "Neonatal mortality by year, all regions."}
neo <- age.window(births, c(0, 1 / 13))
surv <- as.matrix(neo[, c("enter", "exit", "event", "birthdate")])
pall <- perstat(surv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.5), col = "blue", xlab = "Year",
     ylab = "Mortality")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

## By region

```{r neoby, fig.cap = "Neonatal mortality by year  and region."}
usurv <- as.matrix(neo[neo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
pume <-  perstat(usurv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.7), col = "blue", xlab = "Year",
     ylab = "Mortality")
bsurv <- as.matrix(neo[neo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
pjur <-  perstat(bsurv, period = 1900:1951)$intensity
lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
##ssurv <- as.matrix(neo[neo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

# Post-neonatal mortality, an overview

```{r pnneo, fig.cap = "Post-eonatal mortality by year, all regions."}
pneo <- age.window(births, c(1 / 13, 1))
surv <- as.matrix(pneo[, c("enter", "exit", "event", "birthdate")])
pall <- perstat(surv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
     xlab = "Year", ylab = "Mortality")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

## By region

```{r pneoby, fig.cap = "Post-neonatal mortality by year and regions."}
usurv <- as.matrix(pneo[pneo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
pume <-  perstat(usurv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
     xlab = "Year", ylab = "Mortality")
bsurv <- as.matrix(pneo[pneo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
pjur <-  perstat(bsurv, period = 1900:1951)$intensity
lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
##ssurv <- as.matrix(pneo[pneo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

@@

We have two stations, *Umeå* and *Bjuröklubb*.

```{r gettemp}
source("R/tempdat.R")
umetemp <- tempdat("ume", c(1895, 1950))
bjurtemp <- tempdat("bjur", c(1895, 1950))
##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
```

# Make communal

```{r makecommunal}
##source("R/make_comm.R")
lagg <- 3 / 365  ## NOTE!!
## Ume:
bume <- births[births$subreg == "ume", ]
bume <- make.communal(bume, umetemp["heat"], start = 1895, period = 1/52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
bume <- make.communal(bume, umetemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
bume <- make.communal(bume, umetemp["week"], start = 1895, period = 1 / 52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
### Ske:
bskel <- births[births$subreg == "bjur", ]
bskel <- make.communal(bskel, bjurtemp["heat"], start = 1895, period = 1/52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
bskel <- make.communal(bskel, bjurtemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
bskel <- make.communal(bskel, bjurtemp["week"], start = 1895, period = 1 / 52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
##bsten <- births[births$subreg == "sten", ]
##bsten <- make_comm(bsten, stentemp, start = 1899, period = 1/52, lag = lagg)
bume$region <- "ume"
bskel$region <- "ske"
infdat <- rbind(bume, bskel)
infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
                                                   "birthdate", "socStatus", "hisclass", "illeg",
                                                   "parity", "heat", "cold", "week")]
saveRDS(infdat, file = "data/infdat.rds")
```

Done!

>>>>

---
title: "getdata4.Rmd"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skum) # Also loads 'eha'
age.cuts = c(0, 3, 7, 28, 56, 140, 183, 250, 365) / 365
```

# Introduction

*Note* that this is the old getdata4.Rmd, developed after cleaning.

Add *socst* on 13 February 2022.

Added *urban* on 12 February 2022.

Created 6 February 2022 after consultations.

** Run this (and only this) before running "paper4.Rmd"!**

We start with *skum::persons*  and age is to begin with in days,
with stillbirths  *removed* and born alive but dead on birthday at age 0.25 days.

For use of *eha::make.communal*, we need to temporarily change time unit to years.

Further, we eventually skip the use of eha::make.communal in  favor of *match*. Reason: 
eha::make.communal assumes age is measured in *years*. But this is no promise yet.

# Read skum::pers

```{r readpers}
per <- pers[(pers$frsbobtyp == 2) & (pers$df == 0), ]
per$urban <- per$fboort %in% c(274, 1208)
per$frsbobtyp <- per$df <- per$f2id <- per$m2id <- NULL
per$frsbobdatmin <- per$frsbobdatmax <- per$utdatmax <- per$utdatmin <- NULL
per <- per[(per$foddat > 18950000), ]
dn <- per$doddat < 1
per$doddat[dn] <- per$foddat[dn] + 40000
## Fix foddat:
yr <- per$foddat %% 10000 == 0
per$foddat[yr] <- per$foddat[yr] + 101 # As early as possible
mon <- per$foddat %% 100 == 0
per$foddat[mon] <- per$foddat[mon] + 1 # Ditto
## Fix doddat:
yr <- per$doddat %% 10000 == 0
per$doddat[yr] <- per$doddat[yr] + 1231 # As late as possible
mon <- per$doddat %% 100 == 0
per$doddat[mon] <- per$doddat[mon] + 28 # Ditto (guard for February ...)
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
nabir <- is.na(per$birthdate)
per$foddat[nabir] <- per$foddat[nabir] - 2
per$doddat[nabir] <- per$doddat[nabir] - 2
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate))
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
## Fix doddat:
per$doddat[is.na(per$deathdate)] <- per$doddat[is.na(per$deathdate)] - 1
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
##
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
oj <- per$exit < 0
per$doddat[oj] <- per$doddat[oj] + 10000
per$deathdate <-  as.Date(as.character(per$doddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
per$event <- per$exit <= 365
per$exit[per$exit == 0] <- 0.25
per$exit[per$exit > 365] <- 365
```

Now we need *nofrs* for newborn, found in the *observation* file(s), so

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$socst <- bobs$socst[indx]
per$socst[is.na(per$socst)] <- 9
per$socst1 <- factor(per$socst, labels = c("highBusiness", "highOfficial", "farmer", "lowBusiness",
                                          "lowOfficial", "qualWorker", "farmhand", "unqualWorker", "none"))
per$socst <- factor(per$socst, labels = c("highWhiteC", "highWhiteC", "farmer", "lowWhiteC", "lowWhiteC", 
                                           "qualWorker", "farmer", "worker", "none"))
##per$socStatus <- bobs$socStatus[indx]
##per$socBranch <- bobs$socBranch[indx]
per$hisclass <- bobs$hisclass[indx]
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
per$subreg <- factor(per$subreg)
per$enter <- 0
per$sex <- factor(per$kon, labels = c("boy", "girl"))
per$birthdate <- toTime(per$birthdate)
per$illeg <- as.numeric(per$ab == 2)
per$parity <- per$paritet_g
per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
## NEW:
##per$socBranch <- as.character(per$socBranch)
##per$socBranch[is.na(per$socBranch)] <- "none"
##per$socBranch <- factor(per$socBranch)
##per$socStatus <- as.character(per$socStatus)
##per$socStatus[is.na(per$socStatus)] <- "low"
##per$socStatus <- factor(per$socStatus)
per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", "socst",
               "subreg", "hisclass", "urban", "enter", "exit", "event")]
saveRDS(per, file = "data/births4.rds")
```

@@

We have three stations, *Umeå*, *Bjuröklubb*, and *Stensele*.

```{r gettemp}
source("R/tempdat.R")
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##
umetemp <- tempdat("ume", temp_ivl)
bjurtemp <- tempdat("bjur", temp_ivl)

##Changed 5 feb 2022:
##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
source("R/normal.R") # normal = Norsjö - Malå!!
stentemp <- normal(temp_ivl)
## End change

saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
```

# Put on temperature data

Here we must change time unit to *year*.

```{r makecommunal}
per$enter <- per$enter / 365
per$exit <- per$exit / 365
## per$birthdate <- toTime(per$birthdate) #already done!
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##

lagg <- 0 / 365 ##3 / 365  ## NOTE!!
tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
               "week", "year") 
## Ume:
bume <- per[per$subreg == "ume", ]
bume <- make.communal(bume, umetemp[, tempvars], 
                      start = temp_start, period = 1/52, lag = lagg)

### Ske:
bskel <- per[per$subreg == "bjur", ]
bskel <- make.communal(bskel, bjurtemp[, tempvars], 
                       start = temp_start, period = 1/52, lag = lagg)

### Sten:
bsten <- per[per$subreg == "sten", ]
bsten <- make.communal(bsten, stentemp[, tempvars], 
                       start = temp_start, period = 1/52, lag = lagg)

bume$region <- "ume"
bskel$region <- "ske"
bsten$region <- "sten"
infdat <- rbind(bume, bskel, bsten)
infdat$birthmonth <- ceiling((infdat$birthdate - floor(infdat$birthdate)) * 12)
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
                                                   "birthmonth", "period", "socst", "illeg",
                                                   "parity", "hisclass", "urban", tempvars)]
## New addition: (28 Jan 2022)
##infdat <- rc(infdat) # rc from 'skum'
##n <- NROW(infdat)

saveRDS(infdat, file = "data/infdat4.rds") # NOTE: Name change!!!

## Wait with this:
## tabdat4 <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + socBranch + illeg + parity 
```

Note the name change of the data file: *infdat4.rds*! And the the time lag is `r lagg`.
